$p$-adic precision

### 2014-02-20 padicprec.tex
Switch to one-pass and two-pass notation

It is nevertheless not obvious how to implement these ideas in practice
because when we enter in the $i$-th step of the execution of {\tt f},
we have not computed $h_i$ yet and hence are \emph{a priori} not able
to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
and \eqref{item:Hi2} above.
A possible solution to tackle this problem is to proceed in several
stages as follows: 
\begin{enumerate}[(1)]
\item \label{item:smallprec}
for $i$ from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
at small precision (but enough for the second step) together with an
approximation of the Newton polygon of the higher order derivatives;
\item \label{item:determineHi}
for $i$ from $n$ to $1$, we compute $h'_i(x_i)$ and
determine a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
\eqref{item:Hi2};
\item \label{item:finalcomp}
for $i$ from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
and finally outputs $x_n + O\big(f'(x)(H)\big)$.
\end{enumerate}
Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
\cite{hoeven1, hoeven2, berthomieu}), we can reuse in 
Step~\eqref{item:finalcomp} the computations already performed in 
Step~\eqref{item:smallprec}. The ``backward'' method we have just 
presented is then probably not much more expansive than the ``forward'' 
method, although it is certainly much more difficult to implement.

We conclude this section by remarking that the ``backward'' method seems 
to be particularly well suited to computations with lazy $p$-adic. Indeed 
in this setting, a target precision (\emph{i.e.} a precision on the 
output) is fixed and the software determines automatically the precision 
it needs on the input to perform the computation. To do this, it first 
builds the ``skeleton'' of the computation (\emph{i.e.} it determines the 
functions $f_i$ and eventually computes the $x_i$ at small precision when 
branching points occur and it needs to decide which branch it follows) 
and then runs over this skeleton in the reverse direction in order to 
determine (an upper bound of) the needed precision at each step. The 
similarity with our approach is noticeably strong.

@@

Let us illustrate the above discussion by an example: suppose that we 
want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
family of $n$ vectors to its Gram matrix. The differential of $f$ is
clearly never surjective because $f$ takes its values in the subspace
consisting of symmetric matrices. Nevertheless, for all pairs $(i,j)
\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ takes a matrix 
$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
their differentials are generically surjective. Let $M$ be a matrix known
at some finite precision such that 
$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
and/or the backward computation and get $f_{ij}(M)$ together with its 
precision. Putting this together, we get the whole matrix $f(M)$ 
together with a sharp precision datum on each entry.

@@

We conclude this section by briefly remarking that the two strategies designed 
above (\emph{i.e.} forward and backward computations) share some 
similarities with usual floating point arithmetics over the reals. 
Indeed, roughly speaking, in each setting, we begin by choosing a large 
precision, we do all our computations up to this precision understood 
that when we are not sure about some digit, we choose it ``at random'' 
or using good heuristics. The main difference is that, in the 
ultrametric setting, we are able (under some mild hypothesis) to 
quantify the precision we need at each individual step in order to be 
sure that the final result is correct up to the required precision.

>>>>

It is nevertheless not obvious how to implement these ideas in practice
because when we enter in the $i$-th step of the execution of {\tt f},
we have not computed $h_i$ yet and hence are \emph{a priori} not able
to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
and \eqref{item:Hi2} above.
A possible solution to tackle this problem is to proceed in several
stages as follows: 
\begin{enumerate}[(1)]
\item \label{item:smallprec}
for $i$ from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
at small precision (but enough for the second step) together with an
approximation of the Newton polygon of the higher order derivatives;
\item \label{item:determineHi}
for $i$ from $n$ to $1$, we compute $h'_i(x_i)$ and
determine a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
\eqref{item:Hi2};
\item \label{item:finalcomp}
for $i$ from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
and finally outputs $x_n + O\big(f'(x)(H)\big)$.
\end{enumerate}
Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
\cite{hoeven1, hoeven2, berthomieu}), we can reuse in 
Step~\eqref{item:finalcomp} the computations already performed in 
Step~\eqref{item:smallprec}. The two-pass method we have just 
presented is then probably not much more expansive than the one-pass 
method, although it is more difficult to implement.

We conclude this section by remarking that the two-pass method seems 
to be particularly well suited to computations with lazy $p$-adics.
In this setting, a target precision is fixed and the software determines automatically the precision 
it needs on the input to achieve this output precision. To do this, it first 
builds the ``skeleton'' of the computation (\emph{i.e.} it determines the 
functions $f_i$ and eventually computes the $x_i$ at small precision when 
branching points occur and it needs to decide which branch it follows) 
and then runs over this skeleton in the reverse direction in order to 
determine (an upper bound of) the needed precision at each step.

@@

Let us illustrate the above discussion by an example: suppose that we 
want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
family of $n$ vectors to its Gram matrix. The differential of $f$ is
clearly never surjective because $f$ takes its values in the subspace
consisting of symmetric matrices. Nevertheless, for all pairs $(i,j)
\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ takes a matrix 
$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
their differentials are generically surjective. Let $M$ be a matrix known
at some finite precision such that 
$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply a one- or two-pass 
computation and get $f_{ij}(M)$ together with its 
precision. Putting this together, we get the whole matrix $f(M)$ 
together with a sharp precision datum on each entry.

@@

We conclude this section by briefly remarking that the two strategies described 
above share some 
similarities with usual floating point arithmetics over the reals. 
Indeed, roughly speaking, in each setting, we begin by choosing a large 
precision, we do all our computations up to this precision understood 
that when we are not sure about some digit, we choose it ``at random'' 
or using good heuristics. The main difference is that, in the 
ultrametric setting, we are able (under some mild hypothesis) to 
quantify the precision we need at each individual step in order to be 
sure that the final result is correct up to the required precision.
